NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-10-09 Manuscript/manuscriptPS.tex
Update text

\subsection{Counterion binding and interactions between PC and PS headgroups}
Membranes containing PS lipids are always accomppanied with counterions which
modulate electrostatic interactions between lipids and other biomolecules. 
Counterions are also suggested to screen the repulsion between charged lipid headgroups 
in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
The counterion density profiles along membrane normal 
show significant differences between simulation models in both binding affinity
and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
The strongest sodium counterion binding, i.e., the lowest concentrations in bulk water,
is observed in MacRog, Berger and Lipid17/JC simulations.
Slipids, Berger, MacRog, and Lipid17 exhibit a clear single maxima in the
counterion density, but slightly at different distances from the bilayer
center. Other models exhibit more complicated distribution, often including
two local maxima in the density profile.
%CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
%density, while a single maxima is observed in the other models.

Area per lipid is in agreement with experiments \cite{pan14} only
in the MacRog simulation with potassium counterions and in Gromos-CKP models,
while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
The difference cannot be explained only by the electrostatic screening of the headgroup repulsion due to 
counterion binding because CHARMM36, CHARMM36ua and Slipid models give
smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
On the other hand, increasing counterion binding affinity reduces the area per molecule
when comparing between sodium and potassium in MacRog simulations or between
ff99 and jc ions in Lipid 17 simulations.
This suggests that the counterion binding alone cannot explain the low area per molecule
in PS lipid bilayers but also the hydrogen bonding network has to
be taken into account, as also previously suggested based on scattering experiments \cite{petrache04}.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}

Evaluation of counterion binding affinity to charged lipid bilayers
using the changes of headgroup order parameters is difficult because
reference state without ions does not exist, in contrast to previous
studies with zwitterionic membranes and coions \cite{catte16},
see also section \ref{mixtureTOadditionalCIs} in the supplementary information.
However, the counterion binding affinity is
reflected to the increase of POPC headgroup order parameters due to the
increasing amount of negatively charged POPS lipids in the bilayer.
The experimentally measured increase in the order parameters~\cite{scherer87}
is reproduced only in MacRog simulations, while other simulations
predict no change or even decrease in the order parameters (Fig. \ref{HGorderparametersPCvsPS}).
The results can be explained using the electrometer concept \cite{seelig87,scherer89}.
In experiments and MacRog simulations, the increasing amount of negative charge
in the membrane tilts the PC headgroup more parallel to the membrane plane
leading to the larger order parameters. In Berger simulations,
the strong sodium binding surpasses the effect of negatively charged
lipids and the order parameters decrease with the increasing amount of
positive counterions of PS lipid.
In CHARMM36 simulations, the counterion binding neutralizes the negative charge
of lipids at the interface and the headgroup order parameters are not changed
with increasing amount of PS. Therefore, the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) reproduces the correct
response of POPC headgroup to the increasing amount of negatively charged lipids,
indicating that these simulations have realistic surface charge density in membranes
containing PS lipids.
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of PC (left panel) and PS (right panel) headgroup order parameters
    from POPC:POPS mixtures with increasing amount of POPS.
    Experimental results of POPC are taken from Ref. \citenum{scherer87}
    (signs are determined as discussed in \cite{botan15,ollila16}).
    Experimental values for POPS in pure bilayer and in mixture are  
    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
    Since the experimental data of POPS in pure and diluted mixture come from
    different experimental sets (13C NMR in this work and 2H NMR from Ref. \citenum{roux90}),
    the experimental change of the order parameter is less accurate than in typical measurements
    where same technique is used in all conditions, see discussion about qualitative and quantitative 
    accuracy in Ref. \citenum{ollila16}.
    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
    set to the value from POPC:POPS (5:1) mixture.
    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
    with the same value for both order parameters such that the lower order
    parameter value from POPC:POPS (5:1) mixture is at zero
    to correctly illustrate the significant forking.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
\end{figure*}

>>>>

\subsection{Counterion binding and interactions between PC and PS headgroups}
Membranes containing PS lipids are always accomppanied with counterions which
modulate electrostatic interactions between lipids and other biomolecules. 
Counterions are also suggested to screen the repulsion between charged lipid headgroups 
in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
The counterion density profiles along membrane normal 
show significant differences between simulation models in both binding affinity
and distribution of ions in the interface (Fig. \ref{NAdensPOPS}).
The experimental area per lipid (62.7 $\AA^2$) \cite{pan14} 
is reproduced only in Gromos-CKP simulations and in the MacRog simulation
with potassium counterions, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
The counterion binding and concomintant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values
because the MacRog simulation with strongest sodium binding
(the lowest concentrations in bulk water) gives the same area per molecule
as CHARMM36ua simulation with significantly weaker counterion binding
affinity.
%CHARMM36, CHARMM36ua and Slipid models give
%smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
On the other hand, changing counterions from sodium to potassium, having weaker binding
affinity, increase the area per molecule from 53 $\AA^2$ to 63 $\AA^2$ in MacRog simulations.
%with potassium counterions and weaker
%binding affinity give larger area per molecule 
%increasing counterion binding affinity reduces the area per molecule
%when comparing between sodium and potassium in 
%or between
%ff99 and jc ions in Lipid 17 simulations.
In conclusion, the results are in line with the previous study suggesting that the
low area per molecule in PS lipid bilayers originate from the combination
of both counterion binding and hydrogen bonding network between lipid
headgroups \cite{petrache04}.
%, Berger and Lipid17/JC simulations.
%Slipids, Berger, MacRog, and Lipid17 exhibit a clear single maxima in the
%counterion density, but slightly at different distances from the bilayer
%center. Other models exhibit more complicated distribution, often including
%two local maxima in the density profile.
%CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
%density, while a single maxima is observed in the other models. 
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}

Binding of coions to zwitterionic PC lipid bilayers has been previously
evaluated against experiments using the changes of headgroup order parameters
as a function of ion concentration \cite{catte16}. This is less straighforward
for charged lipid bilayers because counterions are always present and the
ion free reference state does not exist
(section \ref{mixtureTOadditionalCIs} in the supplementary information).
However, the counterion binding affinity is
reflected to the increase of POPC headgroup order parameters due to the
increasing amount of negatively charged POPS lipids in the bilayer.
The experimentally measured increase in the order parameters~\cite{scherer87}
is reproduced only in MacRog simulations, while other simulations
predict no change or even decrease in the order parameters (Fig. \ref{HGorderparametersPCvsPS}).
The results can be explained using the electrometer concept \cite{seelig87,scherer89}.
In experiments and MacRog simulations, the increasing amount of negative charge
in the membrane tilts the PC headgroup more parallel to the membrane plane
leading to the larger order parameters. In Berger simulations,
the strong sodium binding surpasses the effect of negatively charged
lipids and the order parameters decrease with the increasing amount of
positive counterions of PS lipid.
In CHARMM36 simulations, the counterion binding neutralizes the negative charge
of lipids at the interface and the headgroup order parameters are not changed
with increasing amount of PS. Therefore, the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) reproduces the correct
response of POPC headgroup to the increasing amount of negatively charged lipids,
indicating that these simulations have realistic surface charge density in membranes
containing PS lipids.
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of PC (left panel) and PS (right panel) headgroup order parameters
    from POPC:POPS mixtures with increasing amount of POPS.
    Experimental results of POPC are taken from Ref. \citenum{scherer87}
    (signs are determined as discussed in \cite{botan15,ollila16}).
    Experimental values for POPS in pure bilayer and in mixture are  
    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
    Since the experimental data of POPS in pure and diluted mixture come from
    different experimental sets (13C NMR in this work and 2H NMR from Ref. \citenum{roux90}),
    the experimental change of the order parameter is less accurate than in typical measurements
    where same technique is used in all conditions, see discussion about qualitative and quantitative 
    accuracy in Ref. \citenum{ollila16}.
    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
    set to the value from POPC:POPS (5:1) mixture.
    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
    with the same value for both order parameters such that the lower order
    parameter value from POPC:POPS (5:1) mixture is at zero
    to correctly illustrate the significant forking.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
\end{figure*}
